---
output: 
  html_document: 
    keep_md: yes
---
# Data Wrangling

```{r, echo=FALSE, results='hide'}
library('knitr')
opts_chunk$set(prompt=TRUE, comment=NA, collapse=TRUE)

library(tidyr, warn.conflicts = FALSE)
library(dplyr, warn.conflicts = FALSE)
```

## Tidy Data

https://cran.r-project.org/web/packages/tidyr/vignettes/tidy-data.html

> Tidy datasets provide a standardized way to link the structure of a dataset (its physical layout) with its semantics (its meaning).

Hadley Wickham provides a philosophical foundation for managing data in his Tidy data paper. We'll work through examples similar to the data shown in the paper.

### Data Structure

The data are from http://www.pricescope.com/wiki/diamond-grading/appendix-ii

```{r}
pricescope_data <- read.delim('pricescope_data.txt', stringsAsFactors = FALSE)
pricescope_data
```

Another way to represent the same data:

```{r}
pricescope_data2 <- read.delim('pricescope_data2.txt', stringsAsFactors = FALSE)
pricescope_data2
```

In tidy data:

1. Each variable forms a column
2. Each observation forms a row
3. Each type of observational unit forms a row

In relational databases, this is Third Normal Form.

```{r, eval=-(1:2), echo = FALSE}
library(tidyr)
library(dplyr)

pricescope_tidy_data <- pricescope_data %>%
  gather(stone, grade, X1:X17) %>%
  mutate(stone = gsub("X", "", stone)) %>%
  mutate(stone = as.numeric(stone)) %>%
  arrange(lab, stone)
pricescope_tidy_data
```

### Tidying Data when Column Headers are Values, Not Variable Names

Our dataset has three variables: lab, stone and grade. To tidy it, we need to _gather_ the values from the headers.

The `gather` function takes three arguments:

1. The name of the _key_ column. That is, the new name of the column that is currently encoded in the headers.
2. The name  of the _value_ column. That is, the new name of the column that is currently held in the data points.
3. The columns to gather.

```{r}
pricescope_tidy_data <- pricescope_data %>%
  gather(stone, grade, X1:X17)
pricescope_tidy_data
```

### The Pipe

The `%>%` pipe helps code readability by piping data from one function to the next.

`f(x, y)` is the same as `x %>% f(y)`

The above code could have been written as:

```{r, results='hide'}
gather(pricescope_data, stone, grade, X1:X17)
```

### Spreading Tidy Data

The `tidyr` package makes it easy to pivot data back into wide format.

```{r}
morley

morley %>%
  spread(Run, Speed)
```

## Dplyr