---
output: 
  html_document: 
    keep_md: yes
---
# Data Wrangling

```{r, echo=FALSE, results='hide'}
library('knitr')
opts_chunk$set(prompt=TRUE, comment=NA, collapse=TRUE)

library(tidyr, warn.conflicts = FALSE)
library(dplyr, warn.conflicts = FALSE)
```

Prerequisite: please copy the model names into the mtcars dataframe:

```{r}
mtcars$model <- rownames(mtcars)
```

## Tidy Data

https://cran.r-project.org/web/packages/tidyr/vignettes/tidy-data.html

> Tidy datasets provide a standardized way to link the structure of a dataset (its physical layout) with its semantics (its meaning).

Hadley Wickham provides a philosophical foundation for managing data in his Tidy data paper. We'll work through examples similar to the data shown in the paper.

### Data Structure

The data are from http://www.pricescope.com/wiki/diamond-grading/appendix-ii

```{r}
pricescope_data <- read.delim('pricescope_data.txt', stringsAsFactors = FALSE)
pricescope_data
```

Another way to represent the same data:

```{r}
pricescope_data2 <- read.delim('pricescope_data2.txt', stringsAsFactors = FALSE)
pricescope_data2
```

In tidy data:

1. Each variable forms a column
2. Each observation forms a row
3. Each type of observational unit forms a table

In relational databases, this is Third Normal Form.

```{r, eval=-(1:2), echo = FALSE}
library(tidyr)
library(dplyr)

pricescope_tidy_data <- pricescope_data %>%
  gather(stone, grade, X1:X17) %>%
  mutate(stone = gsub("X", "", stone)) %>%
  mutate(stone = as.numeric(stone)) %>%
  arrange(lab, stone)
pricescope_tidy_data
```

### Tidying Data when Column Headers are Values, Not Variable Names

Our dataset has three variables: lab, stone and grade. To tidy it, we need to _gather_ the values from the headers.

The `gather` function takes three arguments:

1. The name of the _key_ column. That is, the new name of the column that is currently encoded in the headers.
2. The name  of the _value_ column. That is, the new name of the column that is currently held in the data points.
3. The columns to gather.

```{r}
pricescope_tidy_data <- pricescope_data %>%
  gather(stone, grade, X1:X17)
pricescope_tidy_data
```

### The Pipe

The `%>%` pipe helps code readability by piping data from one function to the next.

`f(x, y)` is the same as `x %>% f(y)`

The above code could have been written as:

```{r, results='hide'}
gather(pricescope_data, stone, grade, X1:X17)
```

### Spreading Tidy Data

The `tidyr` package makes it easy to pivot data back into wide format.

```{r}
morley

morley %>%
  spread(Run, Speed)
```

## Dplyr

Most of this content is from the Introduction to Dplyr at http://rstudio-pubs-static.s3.amazonaws.com/11068_8bc42d6df61341b2bed45e9a9a3bf9f4.html

> When working with data you must:
>
> - Figure out what you want to do.
> - Precisely describe what you want in the form of a computer program.
> - Execute the code.

> dplyr aims to make each of these steps as fast and easy as possible by:
>
> - Elucidating the most common data manipulation operations, so that your options are helpfully constrained when thinking about how to tackle a problem.
> - Providing simple functions that correspond to the most common data manipulation verbs, so that you can easily translate your thoughts into code.
> - Using efficient data storage backends, so that you spend as little time waiting for the computer as possible.

### Five Basic Verbs

Dplyr provides five basic verbs: `filter()`, `arrange()`, `select()`, `mutate()` and `summarise()`

#### Filter()

Filter rows.

```{r}
pricescope_tidy_data %>%
  filter(lab == 'GIA')
```

> Exercise: Using Dplyr syntax, filter the `mtcars` dataset for cars that achieve 25 mpg or greater.

```{r}
mtcars %>%
  filter(mpg >= 25)
```

#### Arrange()

Arrange rows in order.

```{r}
pricescope_tidy_data %>%
  arrange(lab, stone)
```

> Exercise: Order the `mtcars` dataframe by mpg using the `arrange()` function.

```{r}
mtcars %>%
  arrange(mpg)
```

> Exercise: Do the filter (>= 25 mpg) and arrange in one statement

```{r}
mtcars %>%
  filter(mpg >= 25) %>%
  arrange(mpg)
```

#### Select()

Select columns.

```{r}
mtcars %>%
  filter(mpg >= 25) %>%
  arrange(mpg) %>%
  select(model, mpg, cyl, disp)
```

#### Mutate()

Add new columns.

```{r}
mtcars %>%
  mutate(rank = dense_rank(desc(mpg))) %>%
  select(model, mpg, rank)
```

#### Summarize()

Collapse dataframe into a single row.

```{r}
mtcars %>%
  summarize(median_mpg = median(mpg))
```

### Add in some Grouping

In dplyr, you use the group_by() function to describe how to break a dataset down into groups. You can then use the resulting object in the exactly the same functions as above.

What are the mean miles per gallon by cylinder?

```{r}
mtcars %>%
  group_by(cyl) %>%
  summarize(mean_mpg = mean(mpg))
```

> Exercise: Extend the previous statement to find the minimum and maximum mpg for each cylinder group.

```{r}
mtcars %>%
  group_by(cyl) %>%
  summarize(mean_mpg = mean(mpg),
            max_mpg = max(mpg),
            min_mpg = min(mpg))
```

> Exercise: Rank the cars by mpg (with the highest mpg = 1) within each cylinder group. Hint: use `mutate()`. Order the rows by rank and cyl.

Since we are only adding a column, use mutate.

It doesn't work! By default, `dplyr`'s arrange will respect grouping. You must `ungroup()` first.

```{r}
mtcars_ranked <- mtcars %>%
  group_by(cyl) %>%
  mutate(rank = dense_rank(desc(mpg))) %>%
  select(model, cyl, mpg, rank) %>%
  ungroup() %>%
  arrange(rank, cyl)
mtcars_ranked
```

### Window Functions

We've only brushed the surface of what is possible. Try the Window functions vignette. 

https://cran.r-project.org/web/packages/dplyr/vignettes/window-functions.html
