---
output: 
  html_document: 
    keep_md: yes
---

# Plotting with ggplot2

```{r, echo=FALSE, results='hide'}
library('knitr')
opts_chunk$set(prompt=TRUE, comment=NA, collapse=TRUE)

library(tidyr, warn.conflicts = FALSE)
library(dplyr, warn.conflicts = FALSE)
library(ggplot2, warn.conflicts = FALSE)
```

Prerequisite: please copy the model names into the mtcars dataframe:

```{r}
mtcars$model <- rownames(mtcars)
```

## Background

Much of this is from http://opr.princeton.edu/workshops/201501/ggplot2Jan2014DawnKoffman.pdf

`ggplot2` can create most of the statistical graphics from base R, but using a consistent syntax and grammar.

> "attempt to take the good things about base and lattice graphics and improve on them with a strong, underlying model"

Two ways to produce plot objects:

1. ~~`qplot`~~ Do not use! It will only confuse you.
2. `ggplot` is the full implementation of the Grammar of Graphics

The Grammar Defines Components of Graphics

- _data_: in ggplot2, data must be stored as an R data frame
- _geoms_: describe the type of geometric objects that represent data. For example, points, lines, bars, etc.
- _aesthetics_: describe visual characteristics that represent data. For example, position, size, color, shape, transparency, fill
- _facets_: describe how data is split into subsets and displayed as multiple small graphs.
- _stats_: describe statistical transformations that typically summarize data. For example, counts, regression lines, smoothing, etc.
- _coordinate system_: describes the two-dimensional space that data is projected onto. For example, cartesian coordinates, polar coordinates, map projections, etc.
- _scales_: for each aesthetic, describe how visual characteristics are converted to display values. For example, log scales, numeric scales, discrete scales, color scales, etc.

## Building Up a Plot

### A Basic X/Y Plot

We will use the `diamonds` dataset from the ggplot2 package.

```{r, eval=-1}
library(ggplot2)

dim(diamonds)
head(diamonds)
```

Let's examine the relationship of carat weight to diamond price

```{r}
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point()
```

This code shows the following points:

1. The `ggplot` function creates a plot object.
2. Aesthetics map the data from diamonds to the plot
3. Add the `geom_point()` to create the geometric object

Why do we use `+` instead of `%>%`? 

As expected, price increases as carat weight increases. What other patterns do you see in this plot?

> Exercise: Create a plot of the `mtcars` dataset, showing weight (wt) on the x-axis and miles per gallon (mpg) on the y-axis. What is the relationship of the two variables?

```{r}
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point()
```

### Additional Aesthetics

Use the help to understand what additional aesthetics are possible.

```{r, eval=FALSE}
?geom_point
```

Let's try coloring the points based on the clarity grade. How would you do this?

```{r}
ggplot(diamonds, aes(x = carat, y = price, color = clarity)) +
  geom_point()
```

### Understanding Aesthetics vs. Graphical Options

Rember, `aes()` should wrap options _only_ if they change with the data.

Assume you wanted each dot to be blue. In that case, do not use aes for the color.

```{r}
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point(color = "blue")
```

> Exercise: Plot the mtcars dataset with `x = wt, y = mpg`, but make the points vary in size according to the horsepower of the car. What is the relationship?

```{r}
ggplot(mtcars, aes(x = wt, y = mpg, size = hp)) +
  geom_point()
```

### Adding Facets

What if we wanted to examine the relationship of diamond price, with each clarity grade split out into a separate graph?

```{r}
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point() +
  facet_wrap(~ clarity) 
```

Here are some interesting trends. I1 shows a smooth relationship, but as clarity grade increases, there seems to be a larger jump in price at around one carat. How would we examine only the IF stones?

```{r}
diamonds_if_only <- diamonds %>%
  filter(clarity == 'IF')

ggplot(diamonds_if_only, aes(x = carat, y = price)) +
  geom_point() +
  facet_wrap(~ color) 
```

### Stats

> Exercise: create a bar chart of `mtcars` with x = cyl and the height of the bars is the number of cars in that group. Hint: geom_bar() will count up the cars for you.

```{r}
ggplot(mtcars, aes(x = cyl)) +
  geom_bar()
```

Notice that no y aesthetic is required. `stat_bin` will calculate the bar height automatically.

If you wish to create barplots of data without any transformation, you must specify `stat=identity()`

```{r}
ggplot(mtcars, aes(x = model, y = mpg)) +
  geom_bar(stat = "identity")
```

### Coordinates

This is not very easy to read. The model names will be easier to read if we flip the coordinate system.

```{r}
ggplot(mtcars, aes(x = model, y = mpg)) +
  geom_bar(stat = "identity") +
  coord_flip()
```

This plot can also be improved by reordering the models by mpg.

```{r}
ggplot(mtcars, aes(x = reorder(model, mpg), y = mpg)) +
  geom_bar(stat = "identity") +
  coord_flip()
```

### Scales

The final Grammar of Graphics element customizes the scales of the plot.

Imagine we wanted the y-axis to extend to 50, with labels at every five mpg.

```{r}
ggplot(mtcars, aes(x = reorder(model, mpg), y = mpg)) +
  geom_bar(stat = "identity") +
  scale_y_continuous(limits = c(0, 50), breaks = seq(0, 50, by = 5)) +
  coord_flip()
```

Also note the shortcut functions `xlim` and `ylim`.

## A Few Other Geoms

### Smoothing

Nothing says you have to have only one geom!

```{r}
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point() +
  geom_smooth()
```

#### Boxplot

```{r}
ggplot(mtcars, aes(x = factor(cyl), y = mpg)) +
  geom_boxplot() 
```

#### Density and Histograms

```{r}
ggplot(diamonds, aes(x = price)) +
  geom_histogram()
```

```{r}
ggplot(iris, aes(x = Petal.Length, fill = Species)) +
  geom_density(alpha = 0.50) 
```

#### Line

```{r}
ggplot(economics, aes(x = date, y = unemploy)) +
  geom_line()
```

## Customizing the Look and Feel
