---
output: html_document
---
# Introduction

```{r, echo=FALSE, results='hide'}
library('knitr')
opts_chunk$set(prompt=TRUE, comment=NA, collapse=TRUE)
```

## R and R Studio

R is an open-source statistical programming language, good for data analysis, modelling and visualization.

Base R is the foundation of the language. Base R is extended with packages.

Packages are available from CRAN (The Comprehensive R Archive Network). Use the CRAN Task Views to determine which packages meet a statistical requirement you have.

RStudio Desktop is an integrated development environment (IDE) that makes R easier to use.

RStudio sits on top of R but does not replace R.

R, RStudio and packages are updated frequently.

## Getting to Know RStudio

The RStudio Desktop IDE has four panes:

- Source
- Console
- Environment and History
- Files and Output

Use Alt+Shift+k to view the keyboard shortcuts. Especially note Ctrl+1 to move the cursor to Source and Ctrl+2 to move the cursor to the Console.

# Using Base R

## Scalars

Interactively enter numeric expressions to create scalar values. You can also use arithmetic operators.

```{r}
1
1 + 2
5 - 2
5 * 6
24 / 6
```

Logical operators will output a TRUE or FALSE value.

```{r}
1 < 2
1 > 2
2 == 2
2 != 2
2 >= 3
2 <= 3
```

The output of an expression can be saved using the `<-` assignment operator.

```{r}
a <- 5 * 4
a
```

## Vectors

Vectors are an integral part of R. Use the `:` operator to create vector sequences.

```{r}
1:10
10:1
length(1:10)
```

The `c()` function combines values into a vector.

```{r}
c(1, 13, 7, 5)
x <- c(1, 13, 7, 5)
length(x)
sum(x)
mean(x)
```

Notice that the arithmetic operators work element-wise on vectors.

```{r}
a <- 1:4
b <- 5:8

a
b

a + b
a - b
a * b
a / b
```

Logical operators also work element-wise.

```{r}
a <- 1:5
b <- 5:1

a
b

a == b
a > b
a < b
a >= b
a <= b
```

The `[1]` from the output is indicating the first element of the vector. We saw that with scalar values also. In fact, scalar values in R are simply vectors of length 1.

```{r}
1:100
```

If you operate on vectors of unequal lengths, the shorter vector will be recycled.

```{r}
a <- 1:10
b <- c(10, 100)

a
b

a + b # The result is length 10
```

## Vector Indexing

### Positional Indexing

Use the brackets to index into a specific vector element.

```{r}
x <- c(1, 13, 7, 5)
x

x[1]
x[2]
x[3]
x[4]
x[5] # Returns NA
```

The special `NA` value indicates that the result is not available.

You can index with a vector of positions.

```{r}
x

x[3:4]
```

### Logical Indexing

You can index with logical vectors.

```{r}
x

x[c(TRUE, FALSE, TRUE, FALSE)]
```

Recall that a comparison operator will return a logical vector, such as `x > 6`. What will `x[x > 6]` return?

```{r}
x[x > 6]
```

### Indexing for Sorting

Indexing is not always used to subset. It can also be used to order vector elements.

Examine the output of the `order()` function. What does this produce?

```{r}
order(x)
```

Since `order()` is outputting index positions in the sorted order, how would you use this function to sort the `x` vector?

```{r}
x[order(x)]
```

### Indexing for Replacement

You can assign new values to indexed elements.

```{r}
x
x[2] <- 17
x
```

## Data Frames

Data frames are tables built from equal length vectors.

```{r}
mtcars
```

Indexing data frames follows the same principles as vector indexing, except now we're dealing with two dimensions. Indexing is based on `data[rows, columns]`

```{r}
mtcars[1, 3] # Value from row 1 column 3
mtcars[1,  ] # Leave columns blank to return a row with all columns
mtcars[ , 1] # Leave the rows blank to return only a single column  
```

Since column indexing is common, there are shortcuts.

```{r}
mtcars$cyl       # Use the $ operator to obtain a single column as a vector
mtcars[ , 'cyl'] # or refer to the column by name, rather than position
```

Use `str()` to view the structure of a data frame.

```{r}
str(mtcars)
```

Use `names()` to obtain and change column names.

```{r}
names(mtcars)
names(mtcars)[9]
names(mtcars)[9] <- "transmission"
names(mtcars)
```

### Example: Using a logical vector and vectorized function to recode variables

The transmission (am) column holds a `0` if the car has an automatic transmission and a `1` for a manual transmission. Our job is to recode the ones and zeros to text.

```{r}
# Step 1: create a logical vector of the items you wish to change
mtcars$transmission == 1 
# Step 2: Use the vectorized ifelse() function to create the desired vector
ifelse(mtcars$transmission == 1, "manual", "automatic")
# Step 3: Assign the new vector to the data frame
mtcars$transmission <- ifelse(mtcars$transmission == 1, "manual", "automatic")

mtcars
```

> Exercise: Write code to find the mean mpg for the eight cylinder cars only.

```{r}
mean(mtcars[mtcars$cyl == 8, 'mpg'])
```

> Exercise: Order the `mtcars` dataframe by mpg.

```{r}
mtcars[order(mtcars$mpg), ]
```

## Tables

R can easily tabulate data. For example: How many cars have 4, 6 or 8 cylinders?

```{r}
table(mtcars$cyl)
```

Two (or more) dimensional tables are also possible.

```{r}
table(mtcars$transmission, mtcars$cyl)
```

Use `addmargins()` to add the category sums or `prop.table()` to create proportion tables.

```{r}
tbl <- table(mtcars$transmission, mtcars$cyl)
addmargins(tbl)
prop.table(tbl)
```

## Base Graphics

Base Graphics are the plotting functions included with Base R. Other packages such as `ggplot2` extend R's plotting capabilities.

```{r}
plot(mtcars$wt, mtcars$mpg) # As weight increases, mpg decreases
hist(mtcars$mpg)
boxplot(mpg ~ cyl, data = mtcars)
```

## Coding Style

Code is read more often than it is written so proper style, commenting, and formatting is important, especially if you are sharing code with others.

Review Hadley Wickham's R Style Guide for details at http://adv-r.had.co.nz/Style.html.
